Rclean: A Tool for Writing Cleaner, More Transparent Code

### 2020-01-15 joss/paper.Rmd
JOSS paper edits

@@

The software is currently hosted on Github, and we recommend using the
`devtools` library to install directly from the repository
(https://github.com/MKLau/rclean). The package is open-source and
welcomes contributions. Please visit the repository page to report
issues, request features or provide other feedback.

@@

Although, a lot of great work can be done with type of data
provenance, there are limitations. Only using prospective provenance
means that the outcomes of some processes can not be predicted. For
example, if there is a part of a script that is determined by a random
number, the current implementation of prospective provenance can not
predict the path that will be taken through the code. Therefore, the
code cannot be reduced to exclude the pathway that would not be
taken. Such limitations can be overcome with other data provenance
methods. One solution is "retrospective" provenance, which tracks a
computational process as it is executing. Through this active
monitoring process, retrospective provenance can gather specific
information, such as the results relevant to our random number
example. Using retrospective provenance comes at a cost, however, in
that in order to gather it, the script needs to be executed. When
scripts are computationally intensive or contain bugs that stop
execution, then retrospective provenance can not be obtained for part
or all of the code. Some work has already been done in the direction
of implementing retrospective provenance for code cleaning in R (see
http://end-to-end-provenance.github.io).

For example, the existing framework could be extended to support new
provenance capture methods, and there is tremendous potential for the
use of code cleaning in the creation of more robust capsules
[@Pasquier2018].

@@

This work was improved by discussions with ecologists at Harvard
Forest. Much of the work was funded by US National Science Foundation
grant SSI-1450277 for applications of End-to-End Data Provenance.

>>>>

All of these processes rely on the generation of data provenance
[@Carata2014], which is a formal representation of the execution of a
computational process (https://www.w3.org/TR/prov-dm/), to rigorously
determine the the unique computational pathway from inputs to
results. To avoid confusion, note that "data" in this context is used
in a broad sense to include all of the information generated during
computation, not just the data that are collected in a research
project that are used as input to an analysis. Having the formalized,
mathematically rigorous representation that data provenance provides
gaurantees that the analyses that *RClean* conducts are theoretically
sound. Most importantly, it makes it possible to apply a network
search algorithm to determine the minimum and sufficient code needed
to generate the chosen result in the `clean` function.

There are multiple approaches to collecting data provenance, but
*Rclean* uses "prospective" provenance, which analyzes code and uses
language specific information to predict the relationship among
processes and data objects. *Rclean* relies on a library called
*CodeDepends* to gather the prospective provenance for each
script. For more information on the mechanics of the *CodeDepends*
package, see [@Lang2019]. To get an idea of what data provenance is,
take a look at the `code_graph` function. The plot that it generates
is a graphical representation of the prospective provenance generated
for *Rclean*.

```{R prov-graph}

code_graph(script)

@@

The software is currently hosted on Github, and we recommend using the
`devtools` library to install directly from the repository
(https://github.com/ROpenSci/Rclean). The package is open-source and
welcomes contributions. Please visit the repository page to report
issues, request features or provide other feedback.

@@

The existing framework could be extended to support new provenance
capture methods. One possibility is *retrospective provenance*, which
tracks a computational process as it is executing. Through this
active, concurrent monitoring, retrospective provenance can gather
information that static prospective provenance can't. Only using
prospective provenance means that the outcomes of some processes can
not be predicted. For example, if there is a part of a script that is
determined by a random number, the current implementation of
prospective provenance can not predict the path that will be taken
through the code. Therefore, the code cannot be reduced to exclude the
pathway that would not be taken. However, using retrospective provenance comes
at a cost. In order to gather it, the script needs to
be executed. When scripts are computationally intensive or contain
bugs that stop execution, then retrospective provenance can not be
obtained for part or all of the code. Some work has already been done
in the direction of implementing retrospective provenance for code
cleaning in R (see http://end-to-end-provenance.github.io).

To conclude, we hope that `Rclean` makes writing scientifi software
easier. We look forward to feedback and help with extending its
applications, particularly in the area of reproducibility, such as
using code cleaning in the creation of more robust capsules
[@Pasquier2018]. To get involved, report bugs, suggest features,
please visit the project page.

@@

This work was improved by discussions with ecologists at Harvard
Forest and through the helpful review provided by the ROpenSci
community, particuarly Anna Krystalli, Will Landau and Clemens
Schmid. Much of the work was funded by US National Science Foundation
grant SSI-1450277 for applications of End-to-End Data Provenance.
